---
title: "Test"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  error = TRUE
)
set.seed(2020)
```

## Setup

```{r}
library(LSTbook)
```


## Simulation Model


Suppose that some exercise physiologists are researching whether taking a certain vitamin supplement makes people stronger.  Unknown to them, the supplement changes the amount a person can bench-press by:

```{r}
supplement_effect <- 2
```


They do an observational study involving a large number of people:

```{r}
n <- 2000
```

For each person they record whether not the person takes the supplement, and also how much (in pounds) the person can bench-press.  They also happen to record whether the person works out at the Hulk Power Gym or at Planet Fitness.

Here is a simulation-maker for the situation:


```{r}
supplement_sim <- datasim_make(
  .motivation <- rnorm(n, mean = 0, sd = 10),
  workout <- bernoulli(
    n, logodds = .motivation,
    labels = c("Planet Fitness", "Hulk Power")
  ),
  supplement <- bernoulli(
    n, logodds = .motivation,
    labels = c("no", "yes")
    ),
  bench <- 200 + 0.2 * .motivation  + supplement_effect * (supplement == "yes") + rnorm(n, sd = 10)
)
```

The sample:

```{r}
supplement_data <- supplement_sim |>  
  take_sample(n = n, report_hidden = TRUE)
```

Some of the data:

```{r}
#| label: tbl-sample-data
#| tbl-cap: "The first few rows of the sample data."
supplement_data |>  
  head(n = 10) |>  
  knitr::kable()
```


## Intervention


```{r}
supplement_exp <- supplement_sim |>  
  datasim_intervene(
    supplement <- bernoulli(n, prob = 0.5, labels = c("no", "yes")),
    workout <- bernoulli(n, prob = 0.5, labels = c("Planet Fitness", "Hulk Power"))
  )
```

Note that in the above call, `supplement` and `workout` are defined in reverse order to the order they were defined in `supplement_sim`.

Now a sample:

```{r}
supplement_exp_data <- supplement_exp |>  
  take_sample(n = n, report_hidden = TRUE)
```

Look at the data:

```{r}
supplement_exp_data |> 
  head(10) |> 
  knitr::kable()
```

The new columns have the wrong names.

